Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-01-26 main.tex
:pencil: Update introduction

\begin{abstract}
	This paper investigates how a passive \gls{ris} can reshape the \gls{mimo} point-to-point channel in terms of singular values.
	We depart from the widely-adapted diagonal phase shift model to a general \gls{bd} architecture, which provides superior shaping capability thanks to in-group connections between elements.
	An efficient \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in \gls{los} and fully-connected scenarios.
	As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	A low-complexity suboptimal solution based on channel shaping is also proposed, where the decoupled problem is formulated as channel power maximization and solved in closed form iteratively.
	% We then show how channel shaping decouples the design and amounts to a low-complexity suboptimal solution.
	% A low-complexity suboptimal solution decoupling both blocks is also proposed, where the channel shaping subproblem is formulated as channel power maximization and solved in closed form iteratively.
	Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd}-\gls{ris} increases with group size, stemming from stronger subchannel rearrangement and subspace alignment capabilities.
\end{abstract}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that respects customization.
	% serves on demand.
	This is empowered by the recent advances in \gls{ris}, a real-time programmable metasurface of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than relays.
	Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Despite fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}

@@

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	% Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself, allowing one to decouple \gls{ris} and transceiver design.
	Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself.
	This allows one to decouple the \gls{ris}-transceiver design and explore the fundamental limits of channel manipulation.
	For example, diagonal \gls{ris} has been proved useful for improving channel power \cite{Ning2020}, degree of freedom \cite{Ozdogan2020,Li2023h}, condition number \cite{Zheng2022,Huang2023}, and effective rank \cite{ElMossallamy2021,Meng2023} in \gls{mimo}.
	In contrast, \gls{bd}-\gls{ris} can provide a higher channel power but the results are limited to \gls{siso}\footnote{In terms of channel shaping, single-stream \gls{mimo} with given precoder and combiner \cite{Nerini2023} is equivalent to \gls{siso}.}. \cite{Nerini2023} and \gls{miso} \cite{Santamaria2023}.
	While these studies offer promising glimpses into the channel shaping potential of passive \gls{ris}, a comprehensive understanding of its capabilities and limitations is desired, and a universal design framework is missing.
	This paper aims to answer the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

@@

	Second, we propose a \gls{rcg} algorithm for smooth optimization problems of asymmetric\footnote{Although the constraint of symmetric scattering parameters is widely respected in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2023a,Zhou2023,Li2023d,Bartoli2023}, it can be relaxed when the \gls{bd}-\gls{ris} involves asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}. This assumption has been previously adopted in \cite{Li2023b,Li2023c} and symmetry can be enforced on solutions by projection if necessary.} \gls{bd}-\gls{ris} with arbitrary group size.
	Specifically, blockwise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold and evaluated compactly as matrix exponential.
	The proposed method features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, which is used to solve the Pareto singular value problem.
	This is the first paper to tailor an efficient optimization framework that unleashes the design potential of asymmetric \gls{bd}-\gls{ris}.

@@

	Fourth, extensive simulations reveal that the advantage of \gls{bd}-\gls{ris} over diagonal \gls{ris} increases with group size and \gls{mimo} dimensions.
	In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\%, 312\%, 537\% in $1 \times 1$, $4 \times 4$, $16 \times 16$ configuration under Rayleigh fading.
	The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
\end{section}

\begin{section}{\glsentryshort{mimo}-\glsentryshort{pc}}

>>>>

\begin{abstract}
	This paper investigates how a passive \gls{ris} can reshape the \gls{mimo} point-to-point channel in terms of singular values.
	We depart from the widely-adapted diagonal phase shift model to a general \gls{bd} architecture, which provides superior shaping capability thanks to in-group connections between elements.
	An efficient \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in \gls{los} and fully-connected scenarios.
	As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	A low-complexity suboptimal solution based on channel shaping is also proposed, where the decoupled problem is formulated as channel power maximization and solved in closed form iteratively.
	% We then show how channel shaping decouples the design and amounts to a low-complexity suboptimal solution.
	% A low-complexity suboptimal solution decoupling both blocks is also proposed, where the channel shaping subproblem is formulated as channel power maximization and solved in closed form iteratively.
	Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions, stemming from stronger subchannel rearrangement and subspace alignment capabilities.
\end{abstract}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is empowered by the recent advances in \gls{ris}, a real-time programmable metasurface of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than relays.
	Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Despite fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}

@@

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	% Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself, allowing one to decouple \gls{ris} and transceiver design.
	Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself.
	This allows one to decouple the \gls{ris}-transceiver design and explore the fundamental limits of channel manipulation.
	For example, diagonal \gls{ris} has been proved useful for improving channel power \cite{Ning2020}, degree of freedom \cite{Ozdogan2020,Li2023h}, condition number \cite{Zheng2022,Huang2023}, and effective rank \cite{ElMossallamy2021,Meng2023} in \gls{mimo}.
	In contrast, \gls{bd}-\gls{ris} can provide a higher channel power but existing results are limited to \gls{siso}\footnote{In terms of channel shaping, single-stream \gls{mimo} with given precoder and combiner \cite{Nerini2023} is equivalent to \gls{siso}.}. \cite{Nerini2023} and \gls{miso} \cite{Santamaria2023}.
	While these studies offer promising glimpses into the channel shaping potential, a comprehensive understanding of the capabilities and limitations is desired, and a universal design framework is missing.
	This paper aims to answer the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

@@

	Second, we propose a \gls{rcg} algorithm for smooth optimization problems of asymmetric \gls{bd}-\gls{ris} with arbitrary group size.
	Specifically, blockwise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold and evaluated compactly as matrix exponential.
	The proposed method features lower complexity and faster convergence than general manifold optimization \cite{Absil2009,Pan2022d}, and is used to solve the Pareto singular value problem.
	This is the first paper to tailor an efficient optimization framework that unleashes the design potential of asymmetric \gls{bd}-\gls{ris}.

@@

	Fourth, extensive simulations reveal that the advantage of \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions.
	In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\%, 312\%, 537\% over single-connected in $1 \times 1$, $4 \times 4$, $16 \times 16$ configuration under Rayleigh fading.
	The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
	It emphasizes the importance of \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems.
\end{section}

\begin{section}{System Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, receive antennas, respectively.
	The \gls{bd}-\gls{ris} is modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ independent groups.
	Each group contains $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis, we assume the \gls{bd}-\gls{ris} features lossless asymmetric\footnote{While symmetric (impedance) network is often considered in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2023a,Zhou2023,Li2023d,Bartoli2023}, asymmetric passive components (e.g., ring hybrids and branch-line hybrids) may also be reconfigured in real time \cite{Ahn2006}. Asymmetric \gls{bd}-\gls{ris} has been discussed in \cite{Li2023b,Li2023c,Bartoli2023}. Symmetry can be enforced by projection \cite{Fang2023} if necessary.} in-group connections without mutual coupling.
	The overall scattering matrix is thus a block diagonal matrix $\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G)$, where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is a unitary matrix corresponding to group $g \in \mathcal{G} \triangleq \{1,\ldots,G\}$.
